Demographic inference using the SFS with \moments and \demes

### 2025-08-26 chapter/paper.tex
Nicks final revisions to second draft

% local definitions
\usepackage{xspace}
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\reviewercomment}[1]{{\textcolor{red}{REVIEWER: #1}}}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\demes}{\texttt{demes}\xspace}
\newcommand{\demesdraw}{\texttt{demesdraw}\xspace}
\newcommand{\msprime}{\texttt{msprime}\xspace}
\newcommand{\momi}{\texttt{momi}\xspace}
\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
human-Neandertal history using data from two human populations and a
high-coverage ancient sample from the Neandertal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fit. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to desired value(s) in the
demographic model and for which lower and upper bounds may be set. This file
also specifies any relative \texttt{constraints} between pairs of parameters,
to ensure the optimization routine only explores valid model space. Below, we
have included a truncated parameters file -- the full example is available on
GitHub (URL above).

@@

We fit three demographic models to the simulated data. In addition to
reinferring the parameters from the simulated model, we fit two simpler models
to the same data. These had fewer parameters, both omitting the size change
deeper in time in the ancestral population. This is meant to crudely mimic the
scenario that we often face, in which the true history is more complicated than
the parameterized model. It also highlights biases in inferences that can arise
when features of the true history are not included.

@@

\subsection*{Example 2: Inferring human-Neandertal demographic parameters}

This example uses publicly available data to infer a demographic model for two
modern human populations (a northern European and western African population) 
and a Neandertal individual. As shown below, even with just a single sample 
from the Neandertal lineage, we are able to recover key parameters such as the
human-Neandertal divergence time and fraction of ancestry contributed to modern 
Europeans by Neandertals.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal genome published by \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in the high-coverage archaic genome sequences, 
that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many
parameters ($11$ in the example we present), so we constructed it in stages by
first fitting marginal one- and two-population models. When our goal is to infer 
a complex demographic history involving multiple populations, it is helpful to 
start simple, with subsets of the full data, and sequentially add complexity.  
Here, we began by fitting one-population models with ancestral and recent size 
changes for the MSL and GBR populations.  For the MSL model, we allowed three 
epochs with constant effective size in each epoch. We attributed to GBR a sharp 
contraction $\sim$60 kya followed by exponential growth to represent the 
out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
and 29 years as the average generation time. YAML files representing the models 
and parameters in this section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the resulting 
demographic model to get a two-population model that included continuous 
symmetric migration between MSL and GBR and an ancestral size expansion. 
Following this, we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, the Neandertal effective size, and the 
Neandertal-GBR pulse proportion, leaving other parameters fixed. We also 
fitted the divergence time between the sampled Vindija Neandertal lineage and 
the Neandertal lineage which admixed with modern non-African humans. Finding 
that this parameter tended to converge to its lower bound (imposed by the 
estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
of optimization. Rather than fitting it, we fixed the time of the Neandertal-GBR 
pulse to 50 kya. Fixing parameters for which good estimates are available in the 
literature  (see, e.g., \cite{sumer2025earliest}) is often reasonable in the 
first stages of model optimization, and we expected this value to be especially 
poorly constrained due to the small pulse proportion and Neandertal sample size. 
In the second round of optimization on the three-population model, we refit all 
parameters aside from the two event times mentioned above to refine the 
estimates.

After this refitting step, we were left with parameters in good agreement with 
those inferred by prior authors, with a Neandertal-modern human divergence time 
of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

We also fitted a nested model without the Neanderthal-GBR pulse, which had a 
substantially lower likelihood than the pulse model (log-likelihood [LL] 
$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did 
not perform a formal hypothesis test; the null distribution of the 
likelihood-ratio test (LRT) statistic is overdispersed under composite 
likelihood, inflating the rate of false positives and rendering its use 
unjustifiable in this setting. \moments does incorporate a function 
(\texttt{moments.Godambe.LRT\_adjust}) which computes a multiplicative
adjustment, intended to restore the expected null distribution, using 
first-order moment matching \cite{coffman2016computationally}. This method 
requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
test by fitting the nested and pulse models to a large number of bootstrap 
samples and calculating the rate of rejection across replicates-- which has the 
benefit of producing a distribution of LRT statistics, but is time-consuming
and computationally expensive.

>>>>

% local definitions
\usepackage{xspace}
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\reviewercomment}[1]{{\textcolor{red}{REVIEWER: #1}}}
\newcommand{\nwccomment}[1]{{\textcolor{magenta}{NWC: #1}}}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\demes}{\texttt{demes}\xspace}
\newcommand{\demesdraw}{\texttt{demesdraw}\xspace}
\newcommand{\msprime}{\texttt{msprime}\xspace}
\newcommand{\momi}{\texttt{momi}\xspace}
\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

Below, we show two examples of multi-population demographic inference using the
joint SFS. The first example simulates data with known demographic parameters,
which we try to reinfer using both the original model and simpler
misspecified models. In the second example, we infer a relatively simple
Neanderthal-human history using data from two human populations and a
high-coverage ancient sample from the Neandertal lineage. For both of these
examples, we briefly describe data processing and inference setup, emphasizing
the high-level components of the analyses. For detailed information about each
example, including managing data, specifying models, performing inference,
computing confidence intervals and visualizing fits, we refer readers to the
accompanying GitHub repository
(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
is maintained with up-to-date versioning.

In order to optimize parameters in a \demes model, we need a way to specify
which parameters should be fitted. For this, we use a separate YAML-formatted file
to define \texttt{parameters}, each of which points to desired value(s) in the
demographic model and for which lower and upper bounds may be set. This file
also specifies any relative \texttt{constraints} between pairs of parameters,
to ensure the optimization routine only explores valid model space. Below, we
have included a truncated parameters file -- the full example is available on
GitHub (URL above).

@@

We fitted three demographic models to the simulated data. In addition to
reinferring the parameters from the simulated model, we fitted two simpler 
models to the same data. These had fewer parameters, both omitting the size 
change deeper in time in the ancestral population. This is meant to crudely 
mimic the scenario that we often face, in which the true history is more 
complicated than the parameterized model. It also highlights biases in 
inferences that can arise when features of the true history are not included.

@@

\subsection*{Example 2: Inferring Neanderthal-human demographic parameters}

We used publicly available data to infer a demographic model for two
modern human populations (a northern European and western African population) 
and a Neandertal individual. As shown below, even with just a single sample 
from the Neandertal lineage, we are able to recover key parameters such as the
Neanderthal-human divergence time and the fraction of ancestry contributed to 
modern Europeans by Neandertals.

We obtained modern human genome sequences from a recent resequencing of the 
1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
sequence of the Vindija Neandertal genome \cite{prufer2017high}. 
From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
GBR (British from England and Scotland) sequences as our modern human samples.  
We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
were not genotyped in four high-coverage archaic genome sequences, 
lacked high-confidence ancestral state assignments, or fell within 10 kb of 
exonic or promoter regions. This left us with approximately 960 Mb of 
well-characterized and putatively neutrally-evolving sequence.
We estimated the genome-wide SFS using built-in \moments parsing functions.
Details concerning data processing (including our liftover of the Vindija 
sequence to a more recent genome build) can be found in the linked GitHub 
repository.

The model that we wished to fit is relatively simple but still has many
parameters ($11$ in total), so we constructed it in stages by
first fitting marginal one- and two-population models. When we want to infer 
a complex demographic history involving multiple populations, it is helpful to 
start simple by modeling subsets of the full data and sequentially adding
complexity. This procedure gives us initial guesses for parameter values in more
complex models and may help suggest which features are well-supported by the data.
Here, we began by fitting one-population models with ancestral and recent size 
changes for the MSL and GBR populations.  In the MSL model, we allowed three 
epochs of piecewise-constant effective size. We attributed to GBR two 
constant-size epochs followed by a sharp 
contraction and exponential growth beginning $\sim$60 kya to represent the 
out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
and 29 years as the average generation time. YAML files encoding the models 
and parameters in this section can be found in the linked GitHub repository.

We next stitched these one-population models together and refitted the resulting 
demographic model to get a two-population model that included continuous 
symmetric migration between MSL and GBR and an ancestral size expansion. 
Following this, we added a Neandertal branch and fitted the 
Neandertal-human divergence time, the Neandertal effective size, and the 
Neandertal-GBR pulse proportion, leaving all other parameters fixed. We also 
fitted the divergence time between the sampled Vindija Neandertal lineage and 
the Neandertal lineage which admixed with non-African humans. Finding 
that this parameter tended to converge to its lower bound (imposed by the 
estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
of optimization. We also fixed the time of the Neandertal-GBR pulse to 50 kya 
(at the upper range of values inferrred by \cite{sumer2025earliest}). 
Fixing parameters for which good estimates are available in the literature is often 
reasonable in the first stages of model optimization, and we expected this 
value to be especially poorly constrained due to the small pulse proportion and 
Neandertal sample size. In the second round of optimization of the 
three-population model, we refitted all parameters aside from the two times 
mentioned above to refine the estimates.

After this refitting step, we were left with parameters in good agreement with 
those inferred by prior authors, with a Neandertal-human divergence time 
of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
improved through further exploration of model space.

We also fitted a nested model without the Neanderthal-GBR pulse, which had a 
substantially lower likelihood than the pulse model (log-likelihood [LL] 
$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did 
not perform a formal hypothesis test; the null distribution of the 
likelihood-ratio test (LRT) statistic is overdispersed under composite 
likelihood, inflating the rate of false positives and preventing its use 
 in this setting. \moments does incorporate a function 
(\texttt{moments.Godambe.LRT\_adjust()}) which computes a multiplicative
adjustment to the LRT, intended to restore the expected null distribution, using 
first-order moment matching \cite{coffman2016computationally}. This method 
requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
test by fitting the nested and pulse models to a large number of bootstrap 
samples and calculating the rate of rejection across them-- which has the 
benefit of producing a distribution of (nonindependent) LRT statistics, but is
time-consuming and computationally expensive.
